State of the Practice for Lattice Boltzmann Method Software

### 2022-04-13 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to add more background on recommendations

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of a set of 20 questions). From the measurement results, we ranked the
    software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
    consistent with ranking by GitHub stars, suggesting at least a partial
    correlation between the use of best practices and popularity. We find the
    state of the practice to be healthy, with 67\% of the measured packages
    ranking in the top five for at least one quality, the majority of LBM
    generated artifacts corresponding to general recommendations from research
    software developers, common utilization of version control (67\% of
    packages) and adoption of a quasi-agile development process.  Areas of best
    practice to potentially improve include adoption of continuous integration,
    API documentation and enforcement of programming style guides.  We
    interviewed four developers to gain insight into their current pain points.
    Identified challenges include lack of development time, lack of funding, and
    difficulty with ensuring correctness.  Developers are addressing these pain
    points by designing for change, circumventing the oracle problem and
    prioritizing documentation and usability. For future improvements we suggest
    the following: employing linters, conducting rigorous peer reviews, writing
    and submitting more papers on software, growing the number of contributors
    by following current recommendations for open source projects, and
    augmenting the theory manuals to include more requirements relevant
    information.

@@

To focus our efforts, we devised 10 research questions.  The questions are used
to structure the discussion in the paper, so for each research question below we
point to the section that contains our answer.  The questions are inspired by
the research questions that underpin our data collection methodology
\citep{SmithEtAl2021}. We start with identifying the examples of LBM software
where we have access to the source code:

@@

We next wish to assess the representative software to determine how well they
apply current software development best practices.  At this point in the
process, to remove potential user/developer bias, we will base our assessment
only on publicly available artifacts, where artifacts are the documents, scripts
and code that we find in a project's public repository. Example artifacts
include requirements, specifications, user manuals, unit test cases, system
tests, usability tests, build scripts, API (Application Programming Interface)
documentation, READMEs, license documents, process documents, and code.
Following best practices does not guarantee popularity, so we will also compare
our ranking to how the user community itself ranks the identified projects.

@@

To understand the state of the practice we wish to learn the frequency with
which different artifacts appear, the types of development tools used and the
methodologies used for software development.  With this data, we can ask
questions about how LBM software compares to other research software.  That is,
in what ways does LBM follow the trends from other developer communities, and in
which ways is it different?

@@

We investigated the research questions by applying the general methodology
summarized in \citet{SmithEtAl2021}.  The specific application of the
methodology to LBM is reviewed in Section~\ref{methodology}, with the full
details in \citep{Michalski2021}.  The current methodology updates the approach
used in prior assessments of domains like Geographic Information Systems
\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Oceanographic Software \citep{smith2015state}, Seismology software
\citep{SmithEtAl2018}, statistical software for psychology
\citep{SmithEtAl2018_StatSoft} and medical image analysis software
\citep{Dong2021}.

With our methodology we start by identifying 24 LBM software packages.  We then
approximately measure the application of best practices for each package by
filling in a grading template. Compared with our previous methodology (as used
in \citep{SmithEtAl2016} for instance), the new methodology also includes
repository based metrics, such as the number of files, number of lines of code,
percentage of issues that are closed, etc.  With the quantitative data in the
grading template, we rank the software with the Analytic Hierarchy Process
(AHP). After this, as another addition to our previous methodology, we interview
some of the development teams to further understand the status of their
development process. The quantitative and qualitative data is then used to
answer the research questions (Sections~\ref{methodology}
to~\ref{Sec_AddressConcerns}).  Finally, we summarize the threats to validity
(Section~\ref{threats}) and the final conclusions and recommendations
(Section~\ref{conclusion}).

@@

We developed a methodology for evaluating the state of the practice of
research software \citep{SmithEtAl2021}. This methodology can be applied
to a specific domain of research software, which in the current case is LBM
software.  Since we group best practices around conventional software qualities,
the first section below provides definitions for the qualities of interest
(Section~\ref{softwarequalities}), along with information on how we assessed
these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step in the methodology.

\subsection{Software Qualities} \label{softwarequalities}

The following are the software quality definitions used in this exercise, along
with comments regarding their measurement.

@@

	\item[Maintainability] A measure of maintainability is the effort with which
	a software system or component can be modified to correct faults, improve
	performance or other attributes, and satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work,
	maintainability is measured by the quality of the documentation artifacts,
	and the presence of version control and issue tracking. These artifacts can
	greatly decrease the effort needed to modify software. There are many
	documentation artifacts that can improve maintainability, including user and
	developer manuals, specifications, README files, change logs, release notes,
	publications, forums, and instructional websites.

	\item[Modifiability] Modifiability refers to the ease with which stable
	changes can be made to a system and the flexibility of the system to adopt
	such changes \citep{8016712}. We did not directly measure modifiability.
	Instead, developers were asked in interviews if they considered the ease of
	future changes when developing the software packages, specifically changes
	to the structure of the system, modules and code blocks. A follow up
	question is asked if any measures have been taken to ensure the ease of
	future changes.

@@

	\item[Traceability] Traceability refers to the ability to link the software
	implementation and the software artifacts, especially the requirement
	specification \citep{McCallEtAl1977}. Similar to the quality of correctness,
	this requires the availability of some form of requirements specification.
	This quality refers to keeping track of information as it changes forms or
	relates between artifacts. We did not quantitatively measure traceability.
	Instead, developers were asked in interviews how documentation fits into
	their development process.

@@

	\item[Unambiguity] Unambiguity refers to the extent to which two readers
	have similar interpretations when reading software artifacts. In other
	words, artifacts are unambiguous if, and only if, they only have one
	interpretation \citep{IEEE1998}. This state of the practice assessment did
	not quantitatively measure unambiguity, but we did ask developers if they
	believe that their current documentation clearly conveys all necessary
	knowledge to the users.  Follow-up questions ask how they achieved
	unambiguity, or what improvements are needed to achieve it.

@@

The full template consists of 108 questions categorized under nine qualities:
\begin{inparaenum}[(i)]
	\item installability;
	\item correctness and verifiability;
	\item surface reliability;
	\item surface robustness;
	\item surface usability;
	\item maintainability;
	\item reusability;
	\item surface understandability; and,
	\item visibility/transparency. 
\end{inparaenum}

@@

\section{Comparison Between LBM and Research Software for Artifacts}
\label{Sec_CompareArtifacts}

As part of filling in the measurement template the LBM packages were examined
for the presence of artifacts, which were then categorized by frequency. We have
grouped them into common, uncommon, and rare artifacts in
Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
of the software packages. Uncommon artifacts were found in 8 to 15 (30-63\%) of
the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
software packages.  We answer \rqref{RQ_CompareArtifacts} by comparing the
artifacts that we observed in LBM repositories to those observed and recommended
for research software in general. Our comparison may point out areas where some
LBM software packages fall short of current best practices. This is not intended
to be a criticism of any existing packages, especially since in practice not
every project needs to achieve the highest possible quality.  However, rather
than delve into the nuances of which software can justify compromising which
practices we will write our comparison under the ideal assumption that every
project has sufficient resources to match best practices.

\begin{table}[ht!]
\begin{center}
\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
\toprule
\textbf{Common} & \textbf{Uncommon} & \textbf{Rare} \\
\midrule
Authors / Developers List & Change Log / Release Notes & Acknowledgements\\
Getting Started & Citation & API Documentation\\
Installation Guide / Instructions & Contributing & Code Style Guide\\
Issue Tracker & Design Documentation & Developer / Contributor Manual\\
Library Dependency List & Functional Spec. / Notes & FAQ / Forum\\
License & Performance Information & Requirements Spec.\\
List of Related Publications & Test Plan / Script / Cases & Product Roadmap\\
Makefile / Build File & User Manual / Guide & Unistall\\
README File & & Verification and Validation Plan\\
Theory Notes & & Video Guide (including YouTube)\\
Tutorial & & \\
Version Control & & \\
\bottomrule
\end{tabular}
\caption{Artifacts Present in LBM Packages, Classified by Frequency (Sorted Alphabetically)}
\label{artifactspresent}
\end{center}
\end{table}

@@

In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
row, a checkmark in one of the columns means that the corresponding guideline
recommends this artifact.  The last column shows whether the artifact appears in
the measured set of MI software, either not at all (blank), commonly (C),
uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
artifact consistently between guidelines and specific MI software, but the
terminology and the contents of artifacts are not standardized.  The challenge
even exists for the ubiquitous README file.  As illustrated by
\citet{PranaEtAl2018}, the content of README files shows significant variation
between projects.  Although some content is reasonably consistent, with 97\% of
README files contain at least one section describing the `What` of the
repository and 88.5 \% offering some `How` content, other categories are more
variable.  For instance, information on `Contribution`, `Why`', and `Who`,
appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively
\citep{PranaEtAl2018}.

The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
popularity of recommending a given artifact, but it does not imply that the most
commonly recommended artifacts are the most important artifacts. Just because an
artifact is not explicitly recommended in a given guidelines, does not mean that
the artifact is not valued by the guideline authors.  They may have exluded it
because it is out of the scope of their recommendations, or outside of their
experience.  For instance, an artifact related to uninstall is only explicitly
mentioned by \citep{vanGompelEtAl2016}), but other guideline authors would
likely see its value.  They may simply feel that uninstall is implied by
install, or they may have never asked themselves whether separate uninstall
instructions are needed.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{ p{2.5cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1.2cm}p{1cm}p{0.8cm} }
\toprule
~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & LBM\\
\midrule
LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
\midrule
Uninstall &  &  &  &  & & & & \checkmark & & R\\
Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & C\\
Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & C\\
Code of Conduct &  &  &  &  & & & \checkmark & & & \\
Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & R\\
Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & U\\
Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
\midrule
Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
User manual &  &  & \checkmark &  & & & \checkmark & & & U\\
Tutorials &  &  &  &  & & & \checkmark & & & C\\
FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & R\\
\midrule
Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark & C\\
Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & C\\ 
Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
\checkmark & & \checkmark & C\\
\midrule
Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark &  \\
Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
\checkmark& \checkmark & U\\
API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
Test Plan &  & \checkmark &  &  & & \checkmark & & & &  U\\
Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &
\checkmark & \checkmark & \checkmark & U\\
\bottomrule
\end{tabular}
\caption{Comparison of Recommended Artifacts in Software Development Guidelines
to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}
\label{Tbl_Guidelines}
\end{center}
\end{table}

@@

The majority of LBM generated artifacts correspond to general recommendations
from research software developers.  A union of the three columns in
Table~\ref{artifactspresent} mostly corresponds to recommendations made by the
research software community.  Most of the items in Table~\ref{artifactspresent}
are included in guidelines for writing research software, such as the DLR
(German Aerospace Centre) Software Engineering Guidelines
\citep{TobiasEtAl2018}, xSDK (Extreme-scale Scientific Software Development Kit)
Community Package Policies \citep{SmithAndRoscoe2018}, the EURISE (European
Research Infrastructure Software Engineers') Network Technical Reference
\cite{ThielEtAl2020} and CLARIAH (Common Lab Research Infrastructure for the
Arts and Humanities) Guidelines for Software Quality \citep{vanGompelEtAl2016}.
As an example, LBM developers commonly list the external libraries on which the
software depends, as recommended by \citet{BrettEtAl2021},
\citet{SmithAndRoscoe2018} and \citet{vanGompelEtAl2016}. Although in the rare
column, some LBM developers include an FAQ (Frequently Asked Questions) document
as recommended by \citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and
\citep{vanGompelEtAl2016}. LBM developers also follow the ubiquitous
recommendation to have a README file, although, as for other software, the
content of README files is not standardized \citep{PranaEtAl2018}.  Compared to
ocean modelling software \citep{JungEtAl2022}, LBM software pays more
attention to artifacts related to testing.

@@

Although the LBM community participates in most of the practices we found listed
in the general research software guidelines, some recommended practices were not
observed or rarely observed. For instance API documentation was rarely observed
for LBM software, but it is a frequently recommended artifact
\citep{SmithAndRoscoe2018, ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017,
SSI2022, Zadka2018}.  In addition to the items in the last column of
Table~\ref{artifactspresent}, we can add the following community recommended
items that we rarely, if ever, observed:

\begin{itemize}
\item A roadmap showing what is planned for the future \citep{Yo2021,
ThielEtAl2020, vanGompelEtAl2016}.  We did see this information mentioned in
some repos, but we did not observe a specific artifact devoted to this purpose.
\item A code of conduct to explicitly say how developers should treat one
another \citep{Yo2021, ThielEtAl2020}.  We did not observe a policy of this
sort, possibly because the number of developers for each LBM project is small.
\item Programming style guidelines so that programming labels and formatting are
consistent \citep{TobiasEtAl2018, ThielEtAl2020, OrvizEtAl2017, Zadka2018,
vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
some developer guides, but only rarely.
\item Checklists can be used in projects to ensure that best practices are
followed by all developers.  Some examples include checklists merging branches
into master \citep{Brown2015}, checklists for saving and sharing changes to the
project \citep{WilsonEtAl2016}, checklists for new and departing team members
\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
releases \citep{HerouxEtAl2008} and checklists for overall software quality
\citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a checklist for
managing releases, but otherwise they are used only rarely for LBM.
\item Uninstall instructions \citep{vanGompelEtAl2016} were not observed for any
of the LBM projects.
\end{itemize}

The above discussion shows that, taken together, LBM projects fall short of
recommended best practices for research software.  However, LBM software is not
alone in this.  Many, if not most, research projects fall short of best
practices.  Software requirements documentation provides a good example of the
common deviation between recommendations and practice.  Although requirements
documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
SmithAndKoothoor2016}, in practice research software developers often do not
produce a proper requirements specification \citep{HeatonAndCarver2015}.
\citet{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
found that none of the scientists created requirements specifications, unless
regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
showed requirements are the least commonly produced type of documentation for
research software in general. When looking at the pain points for research
software developers, \citet{WieseEtAl2019} found that software requirements and
management is the software engineering discipline that most hurts scientific
developers, accounting for 23\% of the technical problems reported by study
participants.  The lack of support for requirements is likely due to the
perception that up-front requirements are impossible for research software
\citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
requirements is dropped, allowing the requirements to be written iteratively and
incrementally, requirements are feasible \citep{Smith2016}.

>>>>

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of a set of 20 questions). From the measurement results, we ranked the
    software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
    consistent with ranking by GitHub stars, suggesting at least a partial
    correlation between the use of best practices and popularity. We find the
    state of the practice to be healthy, with 67\% of the measured packages
    ranking in the top five for at least one quality, the majority of LBM
    generated artifacts corresponding to general recommendations from research
    software developers, common use of version control (67\% of
    packages) and adoption of a quasi-agile development process.  Areas of best
    practice to potentially improve include adoption of continuous integration,
    API documentation and enforcement of programming style guides.  We
    interviewed four developers to gain insight into their current pain points.
    Identified challenges include lack of development time, lack of funding, and
    difficulty with ensuring correctness.  Developers are addressing these pain
    points by designing for change, circumventing the oracle problem and
    prioritizing documentation and usability. For future improvements we suggest
    the following: employing linters, conducting rigorous peer reviews, writing
    and submitting more papers on software, growing the number of contributors
    by following current recommendations for open source projects, and
    augmenting the theory manuals to include more requirements relevant
    information.

@@

To focus our efforts, we devised 10 research questions.  The questions are used
to structure the discussion in the paper, so for each research question below we
point to the section that contains our answer.  The questions are inspired by
the research questions that underpin our data collection methodology
\citep{SmithEtAl2021, SmithAndMichalski2022}. We start with identifying the
examples of LBM software where we have access to the source code:

@@

We next wish to assess the representative software to determine how well they
apply current software development best practices.  By \emph{best practices} we
mean methods, techniques and tools that are generally believed to improve
software development, like testing and documentation.  As we will discuss in
Section~\ref{empiricalmeasures}, we will categorize our best practices around
software qualities.  At this point in the process, to remove potential
user/developer bias, we will base our assessment only on publicly available
artifacts, where artifacts are the documents, scripts and code that we find in a
project's public repository. Example artifacts include requirements,
specifications, user manuals, unit test cases, system tests, usability tests,
build scripts, API (Application Programming Interface) documentation, READMEs,
license documents, process documents, and code. Following best practices does
not guarantee popularity, so we will also compare our ranking to how the user
community itself ranks the identified projects.

@@

To understand the state of the practice we wish to learn the frequency with
which different artifacts appear, the types of development tools used and the
methodologies used for software development.  With this data, we can ask
questions about how LBM software compares to other research software.  For cases
where data is available, we will compare to other specific domains of research
software, like medical imaging \citep{Dong2021} and ocean modelling
\citep{JungEtAl2022}.  However, most of the literature on the state of the
practice does not focus on specific domains, but rather on all research software
\citep{SmithAndMichalski2022}. We will refer to studies and guidelines that do
not distinguish an application domain as \emph{research software in general}. By
comparing LBM software to research software in general we can determine the ways
LBM follows the trends from other developer communities, and ways in which it is
different.

@@

We investigated the research questions by applying the general methodology
summarized in \citet{SmithEtAl2021, SmithAndMichalski2022}.  The specific
application of the methodology to LBM is reviewed in Section~\ref{methodology},
with the full details in \citep{Michalski2021}.  The current methodology updates
the approach used in prior assessments of domains like Geographic Information
Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Oceanographic Software \citep{smith2015state}, Seismology software
\citep{SmithEtAl2018}, statistical software for psychology
\citep{SmithEtAl2018_StatSoft} and medical image analysis software
\citep{Dong2021}.

With our methodology we start by identifying 24 LBM software packages.  We then
approximately measure the application of best practices for each package by
filling in a grading template. Our analysis points to areas where some LBM
software packages fall short of current best practices.  This is not intended to
be a criticism of any existing packages, especially since not every project is
actively maintained.  Moreover, we recognize that not every project needs to
achieve the highest possible quality.  However, rather than delve into the
nuances of which software can justify compromising which practice we conducted
our comparison under the ideal assumption that every project can have sufficient
resources to match best practices.

Compared with our previous methodology (as used in \citep{SmithEtAl2016} for
instance), measuring the application of best practices now also includes
repository based metrics, such as the number of files, number of lines of code,
percentage of issues that are closed, etc.  With the quantitative data in the
grading template, we rank the software with the Analytic Hierarchy Process
(AHP). After this, as another addition to our previous methodology, we interview
some of the development teams to further understand the status of their
development process. The quantitative and qualitative data is then used to
answer the research questions (Sections~\ref{methodology}
to~\ref{Sec_AddressConcerns}).  Finally, we summarize the threats to validity
(Section~\ref{threats}) and the final conclusions and recommendations
(Section~\ref{conclusion}).

@@

We developed a methodology for evaluating the state of the practice of research
software \citep{SmithEtAl2021, SmithAndMichalski2022}. This methodology can be
applied to a specific domain of research software, which in the current case is
LBM software.  Since we group best practices around conventional software
qualities, the first section below provides definitions for the qualities of
interest (Section~\ref{softwarequalities}), along with information on how to
measure these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step in the
methodology.

\subsection{Measuring Software Qualities} \label{softwarequalities}

To ensure our assessment of best practices is organized and complete, our
measurements are centred around software qualities.  For completeness we used
the the list of software engineering qualities from \citet[p.\
17--33]{GhezziEtAl2003}, excluding interoperability, productivity and
timeliness.  We excluded interoperability because we are focusing on LBM
software alone, not its interaction with other software.  The other two
qualities were excluded because they are process qualities, and we are measuring
product qualities.  To the list from \citet{GhezziEtAl2003} we added
installability, since it is so critical to the success of any practical project,
and reproducibility, since it is so important for research software. Below we
provide information on each quality of interest, including information on how
they are measured.

@@

	\item[Maintainability] A measure of maintainability is the effort with which
	a software system or component can be modified to correct faults, improve
	performance or other attributes, and satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work,
	maintainability is measured by the quality of the documentation artifacts,
	and the presence of version control and issue tracking. These artifacts can
	greatly decrease the effort needed to modify software. There are many
	documentation artifacts that can improve maintainability, including user and
	developer manuals, specifications, README files, change logs, release notes,
	publications, forums, and instructional websites. We also asked developers
	if they considered the ease of future changes when developing the software
	packages.

@@

@@

@@

The full template consists of 108 questions categorized under nine qualities
(Section~\ref{softwarequalities}):
\begin{inparaenum}[(i)]
	\item installability;
	\item correctness and verifiability;
	\item surface reliability;
	\item surface robustness;
	\item surface usability;
	\item maintainability;
	\item reusability;
	\item surface understandability; and,
	\item visibility/transparency. 
\end{inparaenum}

@@

\section{Comparison Between LBM and Other Research Software for Artifacts}
\label{Sec_CompareArtifacts}

As part of filling in the measurement template the LBM packages were examined
for the presence of artifacts, which were then categorized by frequency. We have
grouped the artifacts into common, uncommon, and rare categories in
Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
of the software packages. Uncommon artifacts were found in 8 to 15 (30-63\%) and
rare artifacts were found in 1 to 7 ($<$30\%) of the packages.  We answer
\rqref{RQ_CompareArtifacts} by comparing the artifacts that we observed in LBM
repositories to those observed and recommended for research software in general.
Our comparison may point out areas where some LBM software packages fall short
of current best practices. This is not intended to be a criticism of any
existing packages, especially since in practice not every project needs to
achieve the highest possible quality.  However, rather than delve into the
nuances of which software can justify compromising which practices we will write
our comparison under the ideal assumption that every project has sufficient
resources to match best practices.

\begin{table}[ht!]
\begin{center}
\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
\toprule
\textbf{Common} & \textbf{Uncommon} & \textbf{Rare} \\
\midrule
Authors / Developers List & Change Log / Release Notes & Acknowledgements\\
Getting Started & Citation & API Documentation\\
Installation Guide / Instructions & Contributing & Code Style Guide\\
Issue Tracker & Design Documentation & Developer / Contributor Manual\\
Library Dependency List & Functional Spec. / Notes & FAQ / Forum\\
License & Performance Information & Requirements Spec.\\
List of Related Publications & Test Plan / Script / Cases & Product Roadmap\\
Makefile / Build File & User Manual / Guide & Uninstall\\
README File & & Verification and Validation Plan\\
Theory Notes & & Video Guide (including YouTube)\\
Tutorial & & \\
Version Control & & \\
\bottomrule
\end{tabular}
\caption{Artifacts Present in LBM Packages, Classified by Frequency (Sorted Alphabetically)}
\label{artifactspresent}
\end{center}
\end{table}

@@

In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
row, a checkmark in one of the columns means that the corresponding guideline
recommends this artifact.  The last column shows whether the artifact appears in
the measured set of LBM software, either not at all (blank), commonly (C),
uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
artifact consistently between guidelines and specific MI software, but the
terminology and the contents of artifacts are not standardized.  The challenge
even exists for the ubiquitous README file.  As illustrated by
\citet{PranaEtAl2018}, the content of README files shows significant variation
between projects.  Although some content is reasonably consistent, with 97\% of
README files contain at least one section describing the `What` of the
repository and 88.5 \% offering some `How` content, other categories are more
variable.  For instance, information on `Contribution`, `Why`', and `Who`,
appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively
\citep{PranaEtAl2018}.

The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
popularity of recommending a given artifact, but it does not imply that the most
commonly recommended artifacts are the most important artifacts. Just because an
artifact is not explicitly recommended in a given guideline, does not mean that
the artifact is not valued.  The guideline authors may have excluded it because
it is out of the scope of their recommendations, or outside of their experience.
For instance, an artifact related to uninstall is only explicitly mentioned by
\citep{vanGompelEtAl2016}), but other guideline authors would likely see its
value.  They may simply feel that uninstall is implied by install, or they may
have never asked themselves whether separate uninstall instructions are needed.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{ p{2.5cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1.2cm}p{1cm}p{0.8cm} }
\toprule
~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & LBM\\
\midrule
LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
\midrule
Uninstall &  &  &  &  & & & & \checkmark & & R\\
Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & C\\
Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & C\\
Code of Conduct &  &  &  &  & & & \checkmark & & & \\
Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & R\\
Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & U\\
Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
\midrule
Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
User manual &  &  & \checkmark &  & & & \checkmark & & & U\\
Tutorials &  &  &  &  & & & \checkmark & & & C\\
FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & R\\
\midrule
Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark & C\\
Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & C\\ 
Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
\checkmark & & \checkmark & C\\
\midrule
Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark &  \\
Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
\checkmark& \checkmark & U\\
API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
Test Plan &  & \checkmark &  &  & & \checkmark & & & &  U\\
Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &
\checkmark & \checkmark & \checkmark & U\\
\bottomrule
\end{tabular}
\caption{Comparison of Recommended Artifacts in Software Development Guidelines
to Artifacts in LBM Projects (C for Common, U for Uncommon and R for Rare)}
\label{Tbl_Guidelines}
\end{center}
\end{table}

@@

@@

As Table~\ref{Tbl_Guidelines} shows the LBM community participates in most of
the practices we found listed in the general research software guidelines;
however, some recommended practices were not observed or rarely observed. For
instance API documentation was rarely observed for LBM software, but it is a
frequently recommended artifact \citep{SmithAndRoscoe2018, ThielEtAl2020,
vanGompelEtAl2016, OrvizEtAl2017, SSI2022, Zadka2018}.  In addition to the items
in the last column of Table~\ref{artifactspresent}, we can add the following
community recommended items that we rarely, if ever, observed:

\item \textbf{A roadmap} is considered by some to be a critical document for an
organization to map out the vision and direction for a product offering
\citep{MunchEtAl2019}.  Roadmaps take different forms but they all aim to
address the following: \begin{inparaenum}[i)]
	\item Where are we now?
	\item Where are we going? and
	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}.
Developers can create a roadmap by the following steps:
\begin{inparaenum}[i)]
	\item define and outline a strategic mission and product vision, 
	\item scan the environment, 
	\item revise and distil the product vision to write the product roadmap, 
	\item estimate the product life cycle and evaluate the mix of planned
development efforts. \end{inparaenum} \citep{VahaniittyEtAl2002}.  A product
roadmap has the benefits of providing continuity of purpose, facilitating
stakeholder collaboration and assisting with prioritization \citep{Pichler2012}.
Although we saw some elements of a roadmap in the LBM projects, we did not
observe a specific artifact devoted to this purpose.

\item \textbf{A code of conduct} documents expectations and standards of ethical
behaviour and explicitly states how developers should treat one another
\citep{TouraniEtAl2017}. A code of conduct outlines rules for communication,
establishes enforcement mechanisms for violations and codifies the spirit of a
community, such that anyone can contribute comfortably regardless of, for
example, ethnicity, gender, or sexual orientation \citep{TouraniEtAl2017}. The
top three most popular code's of conduct from \citet{TouraniEtAl2017} are:
\href{https://ubuntu.com/community/code-of-conduct} {Ubuntu Code of Conduct},
\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
{Contributor Covenant}, \href{https://www.djangoproject.com/conduct/} {Django
Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a code of
conduct on the participation of women. Improved inclusivity means a wider pool
of potential contributors. Moreover, a code of conduct facilitates meeting
ethical obligations for developers that abide by a code of ethics, such as the
IEEE Code of Ethics \citep{IEEE1999}, or the Professional Engineers of Ontario
code of ethics \citep[p.\ 23--24]{PEO2021}. For the LBM software, we did not
observe any codes of conduct, possibly because the number of developers for each
LBM project is small and because the idea hasn't yet occurred to any of the
participants.
